TacticToe: Learning to Prove with Tactics

### 2018-08-14 review_response.tex

\documentclass[]{scrartcl}
\usepackage{csquotes}
\MakeOuterQuote{"}

%opening
\title{Response for SCSS-JSC submission 2}
\subtitle{Aligning Concepts across Proof Assistant
Libraries}
\author{Thibault Gauthier \and Cezary Kaliszyk}

\usepackage{blindtext} 
\usepackage{framed}
%\usepackage[parfill]{parskip}

\setlength{\parskip}{1.5mm}
\setlength{\parindent}{0mm}

\begin{document}

\maketitle

We thank the reviewers for the valuable comments and questions. We have 
attempted to correct all
the issues suggested in the reviews and we give further answers to the 
questions raised in this response letter.

\section*{Reviewer 1}

\begin{leftbar}
My main concern about the paper is related to the usability by people that are 
not the authors of the paper. In several places of the paper, the authors 
indicate that some steps of the alignment process are carried out manually, but 
this is not completely clear; so, the authors should clarify this point in 
order to know which steps are conducted manually and which steps are automated. 
Moreover, the output results are huge plain text file, and exploring them is 
really time-consuming (as indicated previously, the authors have already done 
this, but it might be difficult to the same by other users in a different 
context). In addition, the output results must be manually analysed by the user.

Related to the previous point, the authors provide the final data obtained in 
their
experiments; however, they do not provide the programs that they have used. 
Hence, it is not possible to reproduce the authors' results or use their 
algorithms in different libraries and contexts.
\end{leftbar}

We have now provided the source code of all the experiments performed together 
with instructions how to rerun them
to ease the repeatability and reusability of our project.

When it comes to the section on human advice, the purpose is to allow the user 
to intervene in the decision process,
which could allow the algorithms to continue with more precise matches. Indeed, 
such a hybrid method is very
time consuming and was not evaluated in its full extent.
Except for the logical mappings, all presented matches were automatically 
derived.
We have now explained this in more detail in Section 5.4.

\begin{leftbar}Another issue that arose me when reading the paper were some of 
the applications of this work: translating results and conjecturing new 
theorems. Since the alignment of concepts is based on the quantity (and 
quality) of shared properties, how many results do you need to prove in order 
to discover the alignment of concepts? This is probably one drawback of your 
work: the need of big libraries (it is probably not a problem using the already 
developed libraries, but it might be a problem when creating a new library).
\end{leftbar}

Indeed, the proposed approach would not work well with small proof libraries.
However, given the bridges we constructed between all the considered libraries, 
a relatively small development may
map to specific structures in other libraries and allow the import of related 
theorems.

\begin{leftbar}Another point is whether you have noticed that some matchings 
are missed.
I mean, two concepts that were intended to represent the same mathematical 
object, but that are not matched. I guess that you do not have an automatic way 
of discovering those missing matches.
\end{leftbar}

The only proposed method of discovering such missing matches that our algorithm 
supports is through transitive matches.
We have now added some comments about it in section 4.6.

\begin{leftbar}It is a bit strange that you have not included the alignments in 
the same ITP, have you tried that experiment? Have you discovered something 
interesting? As an idea for the future, you could create some kind of alert 
system that inform the user about the matching of concepts in the libraries of 
the ITP to avoid duplications.
\end{leftbar}

We have tried experiments within a single ITP library. For curated libraries 
where no exact duplicates
occur (like the 6 libraries we consider) the alignments found are often 
concepts that are dual. These
include union and intersection of sets in Mizar, or addition and substraction 
in HOL libraries or
closely related concepts like lipschitzian and continuous.

In a work related to your question (presented at CICM Work in Progress last 
year), we tried to find the most likely substitutions (which mappings
should be applied together) and create a large number of conjectured formulas. 
Only 5 percent of them can be reproved. However they have not been too useful 
for current automation methods. This is now also shortly discussed in the paper.

\begin{leftbar}I wonder if you could define some threshold to differentiate the 
optimal matches, approximate matches and singular matches.
\end{leftbar}

Our current method is statistical, therefore the thresholds are estimates 
learned from previous library alignments.
We could also try to prove equivalence (modulo other mappings) of two similar 
concept but this would most of the times fail even for optimal matches because 
of minor differences in the definitions.

\begin{leftbar}
From my point of view, a graph presenting the workflow that is followed in your 
work to
align concepts might be useful for the reader.
\end{leftbar}
We included a graph together with clarifications to the overview of the 
procedure in Section 1.5.

\section*{Reviewer 2}

\subsection*{COMMENTS ABOUT THE RELEVANCE OF THE PROBLEM}

\begin{leftbar}
Automatically generating alignments is an important practical precondition to 
tackle the harder task of transferring knowledge between mathematical libraries 
at the shallow level, i.e. by translating a constant to the corresponding 
constant in the other library (avoiding to obtain two related but not 
identified notions). The problem seems therefore of primary importance.

However, the authors tackle the problem of aligning existing libraries from 
scratch. This needs to be done only once and forall and the automatically found 
result can be throughly corrected and improved by humans. After it is done, the 
most significant scenario, becomes the one where an author is starting a new 
formalization and he needs to pull in results from other libraries on the same 
constants he is interested in. This scenario is rather different for a series 
of reasons: 1) at the beginning of the new development there are usually only a 
small number of theorems proved and this makes it harder to find correct 
alignments; 2) the algorithm to automatically find alignments could be 
restricted to a subset of the system libraries that are already known to be 
potentially interesting.

I would like to hear from the authors what are their expectations on the 
behaviour of their algorithm on the new scenario. Maybe they could also run 
some benchmarks to receive a first impression, e.g. by showing how the matches 
improve when re-running the algorithm after adding one by one the lemmas in one 
theory.
\end{leftbar}

The proposed scenario is indeed quite different, and like the reviewer 
suggests, it would not work very well because of the first reason.
The only way to cope with it would be to automatically explore the theory 
before the alignment algorithm would be strong enough to discover new theorems 
by itself. As for the second reason why the proposed scenario is different, the 
restriction could be automatically found by a first pass of the alignment
algorithm.

Unfortunately for a number of the considered libraries no theory exploration 
mechanisms exist so far. Developing such mechanisms would in particular require 
powerful general purpose automation, which also has been limited in some of the 
systems. We completely agree that combining the presented work with theory 
exploration would be a very interesting line of work, but it would also be a 
large endeavor and has not been considered so far.

\subsection*{COMMENTS ABOUT THE TECHNICAL SOLUTION}
\begin{leftbar}All the choices the authors took seem reasonable. Of course, it 
is possible to think about totally different techniques to automatically 
identify alignments, for example borrowing other tools from AI and machine 
learning. However, being the first to address the problem at this scale, I 
believe they found a great compromise between simpliciy and efficiency. I will 
add later specific minor comments.

However, one remark I want to do now is that the kind of alignments the authors 
want to find should be directly related to the normalization they apply. For 
example, imagine that the authors want to align binary real division from HOL 
to ternary real division in some Coq libraries. Normalization should somehow 
``normalize away'' the proof argument in order to hope to match properties up 
to alpha-conversion. I can easily come up with other similar examples where 
aligning the library of a constructive, predicative type theory with that of an 
higher order logic requires to be more aggressive in the generalization process 
that produces properties. In previsous papers co-authored by the same authors 
certain classes of alignments were identified. I would have liked to see an 
explanation of normalization that is linked directly to those kind of 
alignments.
\end{leftbar}

Indeed, the manually constructed logical mappings are a weakness of our 
procedure.
However we have been able to find the logical mappings rather quickly, and they 
do not need to be extended as proof libraries grow.
It is true that for an algorithm to discover the logical mappings, the actual 
mappings would need to be more flexible.
In this respect, a possible extension of the algorithm could be to allow 
different parts of the term tree to be conceptualized.
However, we expect that the additional flexibility would reduce precision. We 
have included some discussion of this when considering
subterm conceptualizations.

\subsection*{COMMENTS ABOUT THE EVALUATION METHODOLOGY}

\begin{leftbar}
The evaluation of the technique is probably the weakest part of the paper and 
the one that suggests further future improvements. In particular, the algorithm 
only uses positive pre-existing knowledge in order to identify the alignments 
and complex heuristics to try to single out promising alignments and 
dependencies between the alignments. Finally, the responsibility of checking 
the quality of the alignments is totally left to the user.

A significant improvement would be to try to use automatic techniques to verify 
the quality of the alignments based on the generation of new knowledge, both 
positive and negative. For example, suppose that the system detect a possible 
alignment between concept A and B because A and B share a certain number of 
properties with a good score. In order to evaluate the alignment, the system 
could take for example all the properties on A that are not present for B and 
try to conjecture them to hold for B as well. Once the conjecture is done, 
automatic proof checkers and counterexample generators could be used to prove 
or disprove the conjecture. Even if they fail, the conjectures could often 
still be tested on a certain number of randomly generated tests. For example, 
it should not be hard to automatically understand that the alignment between 
natural and integer numbers is not optimal.
\end{leftbar}

This is a very good idea. There are three reasons why we have not been able to 
do this.
First, we do not have powerful enough general-purpose automation for all the 
provers considered.
Second, it is difficult to evaluate a match on its own. It makes little sense 
to map addition to multiplication but not 0 and 1 at the same time. Therefore 
we would need to consider and evaluate structural alignments consisting of a 
coherent set of matches.
Finally, given the number of matches evaluating each match using many  
automated reasoning experiments would be would be prohibitively time-consuming.

The closest to what the reviewer suggests and is currently possible has been 
considered by us in the CICM Work in Progress paper:
focusing on certain structural alignments one can evaluate the newly generated 
conjectures in some of the provers. The results
have been however much weaker than with other applications presented here. This 
is discussed in Section 1.5 and Section 5.

\begin{leftbar}
Coming to non optimal alignments, I would like to understand if the score 
computed by the algorithms is informative or not on the usefulness of the 
alignment. For example, aligning natural numbers with lists is quite useful 
because a natural number is a degenerate case of list (a list of elements of 
the unit type) and every theorem over lists induces a (possibly trivialized) 
theorem on natural numbers. Another example is aligning natural numbers with 
integers: even if the alignment is non optimal, the alignment becomes optimal 
(or almost so) if restricted to subsets of the two types (natural vs positive 
integers). Most other alignments, instead, are really likely to be completely 
useless.
\end{leftbar}

The scores are representative of the number and quality of the property shared 
by the two constants. So for a particular
pair of provers, the actual value of the score is indeed informative about the 
quality of the alignment, and the best
evidence of this are the actual correct alignments being found despite 
thousands of concepts present.

\begin{leftbar}
Did you ask a mathematician expert in the field what he thinks about the 
conjecture?
\end{leftbar}
We did not ask any expert mathematicians in the domain, instead relying of our 
own limited knowledge of the field to make those conjectures about the 
dynamical system.
We tried to prove those conjectures but failed a bit short so we did some 
empirical evaluation at dimension 2 confirming our intuition that the fix point 
we reach is the only stable fix point.

\begin{leftbar}
Minor comments
\end{leftbar}
We added examples describing the normalization, typing levels and logical 
mappings.
We also gave the patterns representing the most frequent properties, and 
commented on the effect of the coherence constraints on the conev
\end{document}

### 2018-08-14 tactictoe_jv.tex
minor
TODO
minor comments from 1st reviewer

\begin{abstract}
We implement a automated tactical prover TacticToe on top of the HOL4 interactive theorem
prover. TacticToe learns from human proofs which mathematical technique is
suitable in each proof situation.
This knowledge is then used in a Monte Carlo tree search algorithm to
explore promising tactic-level proof paths.
On a single CPU, with a time limit of 60 seconds, TacticToe proves 66.4\% of
the 7164 theorems in HOL4's standard library, whereas
E~prover with auto-schedule solves 34.5\%. The success rate rises to 69.0\% by
combining the results of TacticToe and E~prover.
\end{abstract}

@@

\paragraph{Tactics in \holfour.}
A \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals
of the original goal) together
with a \textit{validation function}.
Calling the validation function on a list of \emph{theorems},
corresponding to
the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the
validation
function on the empty list yields the original goal sequent as a theorem.
In this way, tactics implement the plumbing of goal-directed proof
construction, transforming the conjecture by reasoning in a backward manner.
Since validation functions are only important to check the final
proof, we
omit them during the description of the proof search algorithm. We denote by
$t(g)$ the list of goals produced by the application of a tactic $t$ to a goal
$g$.

@@

\begin{itemize}
%The different
%recording steps (parsing, unfolding and replaying) are regrouped into a single
%\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
  search algorithm (Section~\ref{sec:proofsearch}). The MCTS algorithm gives
more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leaves. The policy and
value function are learned
through supervised learning.
\item Proof guidance required by MCTS is given by prediction algorithms
through supervised learning (Section~\ref{s:prediction}).
These predictors are implemented for three kinds of objects: tactics,
theorems, and lists of goals.
\item We introduce an orthogonalization process that eliminates redundant tactics (Section~\ref{sec:ortho}).
\item We introduce a tactic abstraction mechanism
  (Section~\ref{sec:synthesis}), which enables us to create
more general and flexible tactics by dynamically
predicting tactic arguments.
\item The internal ATP \metis is complemented
  by asynchronous calls to \eprover (Section~\ref{sec:atp}) to help \tactictoe
during proof search.
\item Proof recording at the tactic level is made more precise
  (Section~\ref{sec:recording}). We now support pattern
matching constructions and opening SML modules.
\item Evaluation of \tactictoe with a 60 seconds timeout achieves a 66\%
success rate on the standard library. Comparisons between
\tactictoe and \eprover on different type of problems are reported in
Section~\ref{s:experiments}.

@@

We call a goal \textit{unsolved} if it is not in the set
of solved goals. Similarly, we call a node \textit{unsolved} if it does not
belong to the set of solved nodes.
The \textit{open goal} of a node is the first unsolved goal of this node
according to
some preset order. The other unsolved nodes are called \textit{pending nodes}.
If a node is unsolved, then it contains a unique open goal.
During MCTS exploration, we do not explore pending goals of an unsolved node
before its open goal is solved which justifies our terminology. In other words,
our search tree has the following property: if $g$ is a pending node then
$T(g)$ is empty.

@@

\begin{definition}\label{def:prod} (productive tactic)
The application of a tactic $t$ on an open goal $g$ is called
\textit{productive} if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop, i.e., its output does not contain any goals that appear
in the ancestor of the node of the current goal.
\item It is not a redundant step, i.e., there does not exist $t'$ in $T(g)$
such as $t'(g) \subseteq t(g)$.
\end{itemize}

@@

\paragraph{Tactics}
To start with, we build a database of tactics consisting
of goal-tactic pairs recorded from successful tactic applications in human
proofs (recording will be discussed in
Section~\ref{sec:recording}).
During proof search, the recorded goals will be reordered according to their
similarity with a target goal $g$ (usually the open goal of a node).
The recorded pairs and their similarity to $g$ induce an
order on tactics. Intuitively, tactics which have been successful on
goals similar to $g$ should be tried first. Indeed, they are more likely to lead
to a proof.
This predicted tactic order is then transformed into a prior policy of our MCTS
algorithm (see Section~\ref{sec:policy}).
Tactic selection is also used to improve the quality of the database
of tactics during orthogonalization (see Section~\ref{sec:ortho}).

\paragraph{Theorems as Arguments of Tactics}
We first collect a set of theorems for our predictor to select from.
It includes the \holfour theorem database and theorems from the local namespace.
Predicted tactics during proof search and orthogonalization may include
tactics where arguments (lists of theorems) have been erased (see Section
\ref{sec:synthesis}).
We instantiate those tactics by theorems that are syntactically the closest to
the goal as they are more likely to help.
The same algorithm selects suitable premises for ATPs integrated with
\tactictoe (see Section~\ref{sec:atp}).

\paragraph{Lists of Goals as Output of Tactics}
We compile a dataset of tactic outputs during orthogonalization (see
Section~\ref{sec:ortho}).
Some elements of this set will be considered positive examples (see
Section~\ref{sec:evaluation}).
During proof search, given a list of
goals $l$ created by a tactic, we select a set of lists of goals that are most
similar to $l$. The ratio of positive examples in the selection gives us a
prior evaluation for MCTS.

@@

Our prediction algorithms are not always accurate.
Therefore, a selected tactic may
fail, proceed in the wrong direction or even loop.
For that reason, predictions need to be accompanied by a
proof search mechanism that allows for backtracking and
can choose which proof tree to extend next and in which direction.
Our search algorithm is a Monte Carlo tree search~\cite{montecarlo} algorithm
that relies
on a prior evaluation function and a prior policy function. Those priors are
learned through direct supervised learning for the policy and via the data
accumulated during orthogonalization for the evaluation. Therefore, this MCTS
algorithm does not need to rely on roll outs for evaluation. The
policy and evaluation are estimated by the simple supervised k-NN algorithm.
A characteristic of the MCTS algorithm is that it offers a good trade-off
between exploitation and exploration. This means that the algorithm searches
deeper the more promising branches and leaves enough time for the exploration
of less likely alternatives.

@@

\paragraph{Prior policy}\label{sec:policy}
Similarity between goals produced by the tactic predictor are hard
to translate into a prior policy, so we rely solely on the prediction order
to estimate the probability with which each tactic should be tried.
Let $\mathfrak{T}$ be a search tree after a number of MCTS steps.
Let $p$ be a selected node in $\mathfrak{T}$ and $g$ its open goal. We order
the list of $n$ productive tactics already
applied to $g$ by their similarity score. We note the resulting list
$t_0,\ldots,t_{n-1}$.
Let $c_{\mathit{policy}}$ be a constant heuristic optimized during our experiments.
We calculate the prior policy of a child $a_i$ produced by the tactic
$t_i$ by:
\[\mathit{PriorPolicy}(a_i) = (1 - c_{\mathit{policy}})^{i} * c_{\mathit{policy}}\]

@@

In order to speed up the predictions during the search for a proof of a
conjecture $c$, we preselect 500 goal-tactic pairs and 500 theorems and a
larger number of lists of goals induced by the selection of goal-tactic pairs.
Preselected objects are the only objects available to \tactictoe's search
algorithm for proving $c$.

@@

The first idea is to select goal-tactic pairs and theorems by their similarity
with $c$ (i.e. $\mathit{Predictions}^{\mathit{tactic}}_{500} (c)$ and
$\mathit{Predictions}^{\mathit{theorem}}_{500} (c)$).
However, this selection may not be adapted at later stages of
the proof where goals may have drifted significantly from $c$. In order to
anticipate the production of diverse goals, our prediction during preselection
takes dependencies between objects of the same dataset into account.
This dependency relation is asymmetric. Through the relation, each object has
multiple children and at most one parent.
Once a dependency relation is established we can calculate a dependency score
for each object, which is the maximum of its similarity score
and the similarity score of its parent. Finally, the 500 entries with
highest dependency score are preselected in each dataset.

We now give a mathematical definition of the dependency relation for each kind
of object.

\begin{definition}(Dependencies of a goal-tactic pair)\\
Let $\mathfrak{F}$ be the set of recorded goal-tactic pairs.
The dependencies $D^*$ of a goal-tactic pair $(t_0,g_0)$ are
inductively defined by:

\begin{align*}
D_0 &=_{\mathrm{def}} \lbrace (t_0,g_0) \rbrace \\
D_{n+1} &=_{\mathrm{def}} D_n \cup \lbrace (t,g)\in \mathfrak{F}\  |\ \exists
(t',g') \in D_n.\ g \in t'(g') \rbrace  \\
D^* &=_{\mathrm{def}} \bigcup_{i \in \mathbb{N}} D_i\\
\end{align*}
\end{definition}

\begin{definition}(Dependencies of a theorem)\\
Let $\mathfrak{R}$ be the set of recorded theorems.
The dependencies of a theorem $t$ are the set of top-level theorems in
$\mathfrak{R}$ appearing in the proof of $t$. These dependencies are
recorded by tracking the use of top-level theorems through the inference
kernel~\cite{tgck-cpp15}.
\end{definition}

We do preselection also for lists of goals.
To preselect a list of goals $l$ from our database of lists of goals, we
consider the tactic input $g$ that was at the origin of the production of $l$
during orthogonalization. The list $l$ is preselected if and only if $g$
appears in the 500 preselected goal-tactic pairs.
%This has the
%desirable effect that negative and positive examples derived from the same
%input goal are chosen or discarded together.

\section{Orthogonalization}\label{sec:ortho}
Different tactics may transform a single goal in the same way. Exploring such
equivalent paths
is undesirable, as it leads to inefficiency in automated proof search.
To solve this problem, we modify the construction of the tactics database.
Each time a new goal-tactic pair $(t,g)$ is extracted from a tactic proof and
about to be recorded, we consider if there does not already exist a better
tactic for $g$ in our database. To this end, we organize a competition between
the
$k$ closest goal-tactic pairs $\mathit{Predictions}^{\mathit{tactic}}_k(g)$.
In our experiments,
the default orthogonalization radius is $k=20$.
The winner is the tactic that subsumes (see Definition~\ref{def:tacsub}) the
original tactic on $g$ and that
appears in the largest number of goal-tactic pairs in the database.
The winning tactic $w$ is then associated with $g$, producing the pair $(w,g)$
and is stored in the database instead of the original pair $(t,g)$.
As a result, already successful tactics with a large coverage are preferred,
and new tactics are considered only if they provide a different contribution.
We now give a formal definition of the concepts of subsumption and coverage
that are required for expressing the orthogonalization algorithm.

\begin{definition} (Coverage)\\
Let $\mathfrak{T}$ be the database of goal-tactic pairs. We define the
coverage $\textit{\mbox{Coverage}}(t)$ of a tactic $t$ by the number of times
this tactic
appears in
$\mathfrak{T}$. Expressing this in a formula, we get:
\[\textit{\mbox{Coverage}}(t) =_{\mathrm{def}} |\lbrace g\ |\ (t,g) \in
  \mathfrak{T} \rbrace| \]
Intuitively, this notion estimates how general a tactic is by counting the
number of different goals it is useful for.
\end{definition}

\begin{definition} (Goal subsumption)\\
A goal subsumption $\le$ is a partial relation on goals.
Assuming we have a way to estimate the number of steps required to solve a goal,
a % possibly
general and useful subsumption definition is for example:
% could be:
\[g_1 \le g_2  \Leftrightarrow_{\mathrm{def}} g_1 \mbox{ has a proof with a length
shorter
or equal to the proof of }
g_2\]
By default, we however choose for efficiency reasons a minimal subsumption defined by:
\[g_1 \le g_2  \Leftrightarrow_{\mathrm{def}} g_1 \mbox{ is }\alpha\mbox{-equivalent
to } g_2\]
\end{definition}

@@

\subsection{Implementation}
We describe in more detail our implementation of the recording algorithm. It
consists of 4 phases: tactic proof extraction, tactic proof globalization,
tactic unit wrapping, and creation of goal-tactic pairs.

@@

The function \texttt{save} writes the goal-tactic pair to disk increasing the
number of entries in our database of tactics. The function
\texttt{tactic\_of\_sml} interprets the \sml code \texttt{stac}. The tactic is
then applied to the input goal to replicate the original behavior.
After all modified theories are rebuilt, each call to a wrapped tactic in a
tactic proof is recorded as a pair containing its globalized code and its
input goal.

>>>>

\begin{abstract}
We implement an automated tactical prover TacticToe on top of the HOL4 
interactive theorem
prover. TacticToe learns from human proofs which mathematical technique is
suitable in each proof situation.
This knowledge is then used in a Monte Carlo tree search algorithm to
explore promising tactic-level proof paths.
On a single CPU, with a time limit of 60 seconds, TacticToe proves 66.4\% of
the 7164 theorems in HOL4's standard library, whereas
E~prover with auto-schedule solves 34.5\%. The success rate rises to 69.0\% by
combining the results of TacticToe and E~prover.
\end{abstract}

@@

\paragraph{Tactics in \holfour.}
A \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals
of the original goal) together
with a \textit{validation function}.
Calling the validation function on a list of \emph{theorems},
corresponding to
the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the
validation function on the empty list yields the original goal as a theorem.
In this way, tactics implement the plumbing of goal-directed proof
construction, transforming the conjecture by reasoning in a backward manner.
Since validation functions are only important to check the final
proof, we omit them during the description of the proof search algorithm. We 
denote by $t(g)$ the list of goals produced by the application of a tactic $t$ 
to a goal $g$.

@@

\begin{itemize}
%The different
%recording steps (parsing, unfolding and replaying) are regrouped into a single
%\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
  search algorithm (Section~\ref{sec:proofsearch}). The MCTS algorithm gives
more balanced feedback on
the success of each proof step. The policy and
value function are learned
through supervised learning.
\item Proof guidance required by MCTS is given by prediction algorithms
through supervised learning (Section~\ref{s:prediction}).
These predictors are implemented for three kinds of objects: tactics,
theorems, and lists of goals.
\item We introduce an orthogonalization process that eliminates redundant tactics (Section~\ref{sec:ortho}).
\item We introduce a tactic abstraction mechanism
  (Section~\ref{sec:synthesis}), which enables us to create
more general and flexible tactics by dynamically
predicting tactic arguments.
\item The internal ATP \metis is complemented
  by asynchronous calls to \eprover (Section~\ref{sec:atp}) to help \tactictoe
during proof search.
\item Proof recording at the tactic level is made more precise
  (Section~\ref{sec:recording}). We now support pattern
matching constructions and opening SML modules.
\item Evaluation of \tactictoe with a 60 seconds timeout achieves a 66\%
success rate on the standard library. Comparisons between
\tactictoe and \eprover on different type of problems are reported in
Section~\ref{s:experiments}.

@@

We call a goal \textit{unsolved} if it is not in the set
of solved goals. Similarly, we call a node \textit{unsolved} if it does not
belong to the set of solved nodes.
The \textit{open goal} of a node is the first unsolved goal of this node
according to
some preset order. The other unsolved goals are called \textit{pending goals}.
If a node is unsolved, then it contains a unique open goal.
During MCTS exploration, we do not explore pending goals of an unsolved node
before its open goal is solved which justifies our terminology. In other words,
our search tree has the following property: if $g$ is a pending goal then
$T(g)$ is empty.

@@

\begin{definition}\label{def:prod} (productive tactic)
The application of a tactic $t$ on an open goal $g$ is called
\textit{productive} if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop, i.e., its output does not contain any goals that appear
in the ancestors of the node of the current goal.
\item It is not a redundant step, i.e., there does not exist $t'$ in $T(g)$
such as $t'(g) \subseteq t(g)$.
\end{itemize}

@@

\paragraph{Tactics}
To start with, we build a database of tactics
$\mathbb{D}_{\mathit{tactic}}$ consisting
of tactic-goal pairs recorded from successful tactic applications in human
proofs (recording will be discussed in
Section~\ref{sec:recording}).
During proof search, the recorded goals will be reordered according to their
similarity with a target goal $g$ (usually the open goal of a node).
The recorded pairs and their similarity to $g$ induce an
order on tactics. Intuitively, tactics which have been successful on
goals similar to $g$ should be tried first. Indeed, they are more likely to lead
to a proof.
This predicted tactic order is then transformed into a prior policy of our MCTS
algorithm (see Section~\ref{sec:policy}).
Tactic selection is also used to improve the quality of the database
of tactics during orthogonalization (see Section~\ref{sec:ortho}).

\paragraph{Theorems as Arguments of Tactics}
We first collect a database of theorems $\mathbb{D}_{\mathit{theorem}}$ for our 
predictor to select from.
It includes the \holfour theorem database and theorems from the local namespace.
Predicted tactics during proof search and orthogonalization may include
tactics where arguments (lists of theorems) have been erased (see Section
\ref{sec:synthesis}).
We instantiate those tactics by theorems that are syntactically the closest to
the goal as they are more likely to help.
The same algorithm selects suitable premises for ATPs integrated with
\tactictoe (see Section~\ref{sec:atp}).

\paragraph{Lists of Goals as Output of Tactics}
We compile a dataset of tactic outputs $\mathbb{D}_{\mathit{goallist}}$ during 
orthogonalization (see
Section~\ref{sec:ortho}).
Some elements of this set will be considered positive examples (see
Section~\ref{sec:evaluation}).
During proof search, given a list of
goals $l$ created by a tactic, we select a set of lists of goals that are most
similar to $l$. The ratio of positive examples in the selection gives us a
prior evaluation for MCTS.

@@

Our prediction algorithms are not always accurate.
Therefore, a selected tactic may
fail, proceed in the wrong direction or even loop.
For that reason, predictions need to be accompanied by a
proof search mechanism that allows for backtracking and
can choose which proof tree to extend next and in which direction.
Our search algorithm is a Monte Carlo tree search~\cite{montecarlo} algorithm
that relies on a prior evaluation function and a prior policy function. Those 
priors are learned through direct supervised learning for the policy and via 
the data accumulated during orthogonalization for the evaluation. The
policy and evaluation are estimated by the simple supervised k-NN algorithm.
A characteristic of the MCTS algorithm is that it offers a good trade-off
between exploitation and exploration. This means that the algorithm searches
deeper the more promising branches and leaves enough time for the exploration
of less likely alternatives.

@@

\paragraph{Prior policy}\label{sec:policy}
Let $\mathfrak{T}$ be a search tree after a number of MCTS steps.
Let $n$ be the number of predictions returned by the prior policy (500 in our 
experiments).
Let $p$ be a selected parent node in $\mathfrak{T}$ and $g$ its open goal. We 
order the list of $n$ productive tactics already
applied to $g$ by their similarity score. Indeed, a tactic is likely to 
be useful to solve $g$ if it has been applied on a goal similar to 
$g$ in a previous proofs. 
We note the resulting list
$t_0,\ldots,t_{n-1}$.
Let $c_{\mathit{policy}}$ be a constant heuristic optimized during our experiments.
We calculate the prior policy of a child $a_i$ produced by the tactic
$t_i$ by:
\[\mathit{PriorPolicy}(a_i) = (1 - c_{\mathit{policy}})^{i} * c_{\mathit{policy}}\]

@@

In order to speed up the predictions during the search for a proof of a
conjecture $c$, we preselect 500 tactic-goal pairs and 500 theorems and a
larger number of lists of goals induced by the selection of tactic-goal pairs.
Preselected objects are the only objects available to \tactictoe's search
algorithm for proving $c$.

@@

The first idea is to select tactic-goal pairs and theorems by their similarity
with $c$ (i.e. $\mathit{Predictions}^{\mathit{tactic}}_{500} (c)$ and
$\mathit{Predictions}^{\mathit{theorem}}_{500} (c)$).
However, this selection may not be adapted at later stages of
the proof where goals may have drifted significantly from $c$. In order to
anticipate the production of diverse goals, our prediction during preselection
takes dependencies between objects of the same dataset into account.
This dependency relation is asymmetric. Through the relation, each object has
multiple children and at most one parent.
Once a dependency relation is established (see Definition~\ref{def:tacdep} and 
Definition~\ref{def:thmdep}), we can calculate a dependency score
for each object, which is the maximum of its similarity score
and the similarity score of its parent. Finally, the 500 entries with
highest dependency score are preselected in each dataset.

\begin{definition}\label{def:tacdep}(Dependencies of a tactic-goal pair)\\
Let $\mathbb{D}_{\mathit{tactic}}$ be the database of tactic-goal pairs.
The dependencies $D^*$ of a tactic-goal pair $(t_0,g_0)$ are
inductively defined by:

\begin{align*}
D_0 &=_{\mathrm{def}} \lbrace (t_0,g_0) \rbrace \\
D_{n+1} &=_{\mathrm{def}} D_n \cup \lbrace (t,g)\in 
\mathbb{D}_{\mathit{tactic}}\  |\ \exists
(t',g') \in D_n.\ g \in t'(g') \rbrace  \\
D^* &=_{\mathrm{def}} \bigcup_{i \in \mathbb{N}} D_i\\
\end{align*}
\end{definition}

\begin{definition}\label{def:thmdep}(Dependencies of a theorem)\\
The dependencies of a theorem $t$ are the set of top-level theorems in
the database of theorems $\mathbb{D}_{\mathit{theorem}}$ appearing in the proof 
of $t$. These 
dependencies are recorded by tracking the use of top-level theorems through the 
inferences of the
kernel~\cite{tgck-cpp15}.
\end{definition}

We do preselection also for lists of goals.
To preselect a list of goals $l$ from our database of lists of goals 
$\mathbb{D}_{\mathit{goallist}}$, we
consider the tactic input $g$ that was at the origin of the production of $l$
during orthogonalization. The list $l$ is preselected if and only if $g$
appears in the 500 preselected tactic-goal pairs.
%This has the
%desirable effect that negative and positive examples derived from the same
%input goal are chosen or discarded together.

\section{Orthogonalization}\label{sec:ortho}
Different tactics may transform a single goal in the same way. Exploring such
equivalent paths
is undesirable, as it leads to inefficiency in automated proof search.
To solve this problem, we modify the construction of the tactics database.
Each time a new tactic-goal pair $(t,g)$ is extracted from a tactic proof and
about to be recorded, we consider if there does not already exist a better
tactic for $g$ in our database. To this end, we organize a competition between
the
$k$ closest tactic-goal pairs $\mathit{Predictions}^{\mathit{tactic}}_k(g)$.
In our experiments,
the default orthogonalization radius is $k=20$.
The winner is the tactic that subsumes (see Definition~\ref{def:tacsub}) the
original tactic on $g$ and that
appears in the largest number of tactic-goal pairs in the database.
The winning tactic $w$ is then associated with $g$, producing the pair $(w,g)$
and is stored in the database instead of the original pair $(t,g)$.
As a result, already successful tactics with a large coverage are preferred,
and new tactics are considered only if they provide a different contribution.
We now give a formal definition of the concepts of subsumption and coverage
that are required for expressing the orthogonalization algorithm.

\begin{definition} (Coverage)\\
Let $\mathfrak{T}$ be the database of tactic-goal pairs. We define the
coverage $\textit{\mbox{Coverage}}(t)$ of a tactic $t$ by the number of times
this tactic
appears in
$\mathfrak{T}$. Expressing this in a formula, we get:
\[\textit{\mbox{Coverage}}(t) =_{\mathrm{def}} |\lbrace g\ |\ (t,g) \in
  \mathfrak{T} \rbrace| \]
Intuitively, this notion estimates how general a tactic is by counting the
number of different goals it is useful for.
\end{definition}

\begin{definition} (Goal subsumption)\\
A goal subsumption $\le$ is a partial relation on goals.
Assuming we have a way to estimate the number of steps required to solve a goal,
a % possibly
general and useful subsumption definition is for example:
% could be:
\[g_1 \le g_2  \Leftrightarrow_{\mathrm{def}} g_1 \mbox{ has a proof with a length
shorter
or equal to the proof of }
g_2\]
In this example, the definition of the 
length of a proof is not specified, different definitions may produce different 
subsumptions.
By default, we however choose for efficiency reasons a minimal subsumption defined by:
\[g_1 \le g_2  \Leftrightarrow_{\mathrm{def}} g_1 \mbox{ is }\alpha\mbox{-equivalent
to } g_2\]
\end{definition}

@@

\subsection{Implementation}
We describe in more detail our implementation of the recording algorithm. It
consists of 4 phases: tactic proof extraction, tactic proof globalization,
tactic unit wrapping, and creation of tactic-goal pairs.

@@

The function \texttt{save} writes the tactic-goal pair to disk increasing the
number of entries in our database of tactics. The function
\texttt{tactic\_of\_sml} interprets the \sml code \texttt{stac}. The tactic is
then applied to the input goal to replicate the original behavior.
After all modified theories are rebuilt, each call to a wrapped tactic in a
tactic proof is recorded as a pair containing its globalized code and its
input goal.
